---
title: "CW_cattle-impacts_analysis"
output: html_document
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmmTMB)
library(lubridate)
library(ordinal)
```

## Reading in Data

To Do: decide how to consider streams beyond search distance from plot -- arbitrary large distance or use GIS to get spatial analysis distance
Or consider taking out the distance to water component, since so many plots have 0s with water in the plot

```{r data-cleaning}
cattle_impact_full <- read.csv("Cleaned_Capstone_Data.csv") %>% # reading in csv
  rename_with(tolower) %>%
  rename(  # cleaning up and simplifying column names
    plot = plot._number,
    watershed = subwatrshed_location, 
    hgm = hydrogeomorphic_classification
  ) %>%
  mutate(  # Editing data in our columns for cleaning (use with caution!)
    watershed = str_trim(watershed), # remove whitespace
    grazing_intensity = str_trim(grazing_intensity),  
    # factoring grazing intensity
    grazing_intensity = factor(case_when(grazing_intensity == "Light" ~ "Low", 
                                         TRUE ~ grazing_intensity), 
                               levels = c("None", "Low", "Moderate", "High")), 
    closest_water_distance = as.numeric(closest_water_distance), # formatting Closest_Water_Distance column as a number
    date = as.Date(date, format = "%m/%d/%Y"), # Reformats the Date column into a format which R can interpret as dates correctly
    date = case_when(is.na(date) ~ as.Date("09/09/2025", format = "%m/%d/%Y"),
                     TRUE ~ date) # fixing three dates that came in with the wrong format
  ) %>%
  filter(!is.na(plot)) # removing one blank row


cattle_impact <- cattle_impact_full  %>%
  #  select(date, plot, watershed, hgm, percent_canopy:manure_count, 
  #         closest_water_distance, water_type, restoration_location,
  #         stream_restoration_type:cattle_trail_crossing_or_bedding_area) %>%  # selecting relevant columns
  mutate(
    
    cattle_impact_on_stream = str_trim(cattle_impact_on_stream),
    
    hgm = str_trim(hgm),
    
    hgm_2 = case_when(
      hgm %in% c("Riparian_Mid_Gradient", "Riparian_High_Gradient") ~ "Riparian",
      hgm %in% c("Subsurface_Low_Gradient", "Subsurface_Mid_Gradient", "Subsurface_High_Gradient") ~ "Subsurface",
      TRUE ~ hgm),
    
    restoration_location = replace_na(str_trim(restoration_location), "None"),
    
    restoration_location_2 = case_when(
      restoration_location %in% c("Both", "Inside", "Downstream") ~ "in_or_downstream", 
      restoration_location == "Upstream" ~ "upstream",
      TRUE ~ "none"),
    
    stream_restoration_type = replace_na(str_trim(stream_restoration_type), "None"),
    
    hoof_punch_present = case_when(
      percent_hoof_punch > 0 ~ "yes",
      TRUE ~ "no"),
    
    dc_present = percent_dc > 0, # plots with or without Darlingtonia as a fen indicator
    
    # proportions instead of percents for beta distribution models
    percent_hoof_punch_2 = 
      case_when(
        percent_hoof_punch == 0 ~ 0.001,
        TRUE ~ percent_hoof_punch/100),
    percent_trample_2 = 
      case_when(
        percent_trample == 0 ~ 0.001,
        TRUE ~ percent_trample/100),
    percent_dc_grazed_2 = case_when(
      percent_dc_grazed == 0 ~ 0.001,
      TRUE ~ percent_dc_grazed/100),
    
    # If stream distance wasn't recorded it was > 50m, so putting that in as our minimum for now
    closest_water_distance_2 = case_when(
      is.na(closest_water_distance) ~ 50, 
      TRUE ~ closest_water_distance
    ),
    
    # Binary version of grazing intensity
    grazing_high = case_when(
      grazing_intensity %in% c("None", "Low") ~ 0,
      grazing_intensity %in% c("Moderate", "High") ~ 1,
      TRUE ~ NA
    ),
    
    # splitting stream impacts into separate columns
    stream_impacts = tolower(stream_impacts),
    stream_impacts = replace_na(stream_impacts, ""), # some sites with no stream impacts imported as na rather than blank or none
    stream_incised = str_detect(stream_impacts, "inci"),
    stream_sloughing = str_detect(stream_impacts, "slough"),
    stream_grazing = str_detect(stream_impacts, "graz"),
    stream_trample = str_detect(stream_impacts, "tramp"),
    stream_hoof_punch = str_detect(stream_impacts, "hoof"),
    stream_manure = str_detect(stream_impacts, "manur"),
    stream_trail = str_detect(stream_impacts, "trail|cross"),
    
    
    # adding early vs late visit timing information
    timing = case_when(
      date > "2025-09-01" ~ "Late",
      date <= "2025-09-01" ~ "Early",
      TRUE ~ "NA"
    )
  ) 


# early vist (all)
cattle_early <- cattle_impact %>%
  filter(timing == "Early")

# late visit (subset)
cattle_late <- cattle_impact %>%
  filter(timing == "Late")

# repeated plots both visits
late_plots <- cattle_impact %>%
  filter(timing == "Late") %>%
  pull(plot)

cattle_impact_repeat <- cattle_impact %>%
  filter(plot %in% late_plots) %>%
  arrange(plot)


```

# 0: Summaries of cattle impacts

Want to also add information like percent water, canopy, water within 5m and within 10 m?
Would that go into the same table or a separate one?

Also do we want the late summaries here or are they just distracting?

What if the table design goes the other way, with columns under CM and RF for proportion of plots present, median, mean +- sd
and the rows are the different components of the data? 
```{r cattle-summaries}

plot_summary <- cattle_impact %>%
  group_by(timing, watershed) %>%
  summarise(
    n = n(), 
    canopy_mean = mean(percent_canopy), 
    
  )

cattle_summary <- cattle_impact %>% 
  group_by(timing, watershed) %>%
  summarise(
    n = n(),
    grazing_present = sum(grazing_intensity %in% c("Low", "Moderate", "High"))/n, # proportions of plots
    high_grazing_present = sum(grazing_intensity == "High")/n, 
    trampling_present = sum(percent_trample > 0)/n,  
    trampling_median_cover = median(percent_trample),  # means or medians or both?
    hoof_punch_present = sum(percent_hoof_punch > 0)/n,
    hoof_punch_median_cover = median(percent_hoof_punch), 
    manure_present = sum(manure_count > 0, na.rm = TRUE)/n,   # one plot missing manure count
    manure_median_count = median(manure_count, na.rm = TRUE)
    
    # maybe better not as sd but as range/quartiles? Since sd > median for most of these, asymmetrical values
    # in that case is a table most useful or a chart? 
    #    trampling_sd = sd(percent_trample),
    
    #    hoof_punch_sd = sd(percent_hoof_punch),
    #    manure_sd = sd(manure_count, na.rm = TRUE)
  )

cattle_summary


```




# 1: Predictors of cattle impacts
How to model grazing intensity? 

```{r cattle-impacts}

# hoof punch with beta distribution
glm.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ closest_water_distance_2 + 
                            percent_water + 
                            percent_herb + 
                            percent_canopy + 
                            watershed +
                            (1|hgm_2),
                          data = cattle_early,
                          family = beta_family(link = "logit"))
summary(glm.hoof.punch)

# trampling with beta distribution
glm.trample <- glmmTMB(percent_trample_2 ~ closest_water_distance_2 + 
                         percent_water + 
                         percent_herb + 
                         percent_canopy + 
                         watershed +
                         (1|hgm_2),
                       data = cattle_early,
                       family = beta_family(link = "logit"))
summary(glm.trample)


# manure with poisson distribution
glm.manure <- glmmTMB(manure_count ~ closest_water_distance_2 + 
                        percent_water + 
                        percent_herb + 
                        percent_canopy +
                        watershed +
                        (1|hgm_2),
                      data = cattle_early, 
                      family = poisson(link = "log"))
summary(glm.manure)


# Ordinal grazing intensity data -- cumulative link models using the 'ordinal' package
# Not sure how to interpret this one yet
# could group to med-high and low-none if needed
m.grazing <- clmm(grazing_intensity ~ closest_water_distance_2 + 
                    percent_water + 
                    percent_herb + 
                    percent_canopy +
                    (1|hgm_2),
                  data = cattle_early) 
summary(m.grazing)

# none-low vs moderate-high grazing using a binomial model
m.grazing.2 <- glmmTMB(grazing_high ~ closest_water_distance_2 + 
                         percent_water + 
                         percent_herb + 
                         percent_canopy +
                         watershed +
                         (1|hgm_2),
                       data = cattle_early, 
                       family = binomial(link = "logit"))
summary(m.grazing.2)


```


# 2: Connecting cattle impacts to Darlingtonia
How do cattle impacts correlate with Darlingtonia presence (as a fen indicator)? Where are the biggest direct impacts on Darlingtonia (hoof punch and grazing of Darlingtonia)?
```{r dc}

# hoof punch is higher in sites with dc present (which are the wet sites)
dc.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ dc_present + 
                           percent_dc +
                           (1|hgm_2),
                         data = cattle_early,
                         family = beta_family(link = "logit"))
summary(dc.hoof.punch)


# trample and manure don't differ by dc presence
dc.trample <- glmmTMB(percent_trample_2 ~ dc_present + 
                        percent_dc +
                        (1|hgm_2),
                      data = cattle_early,
                      family = beta_family(link = "logit"))
summary(dc.trample)

dc.manure <- glmmTMB(manure_count ~ dc_present + 
                       percent_dc +
                       (1|hgm_2),
                     data = cattle_early,
                     family = poisson(link = "log"))
summary(dc.manure)



# not sure if this is useful -- overall grazing intensity
dc.grazing <- clm(grazing_intensity ~ dc_present + 
                       percent_dc ,
                  data = cattle_early) 
summary(dc.grazing)

dc.present <- c(TRUE, FALSE)
percent.dc <- c(10, 0)
newdata <- tibble(dc_present = dc.present, percent_dc = percent.dc)

predict(dc.grazing, newdata)



# filter to sites where dc is present -- 35 sites
# to look at where percent grazing of dc is highest
# and hoof punch in dc sites is highest
# no clear drivers of these patterns
cattle_dc <- cattle_early %>% filter(dc_present == TRUE)

m.dc.grazed <- glmmTMB(percent_dc_grazed_2 ~ 
                         closest_water_distance_2 + 
                         percent_dc +
                         percent_water + 
                         percent_herb + 
                         percent_canopy +
                         (1|hgm_2),
                       data = cattle_dc)
summary(m.dc.grazed)

m.dc.hoofpunch <- glmmTMB(percent_hoof_punch ~ 
                         closest_water_distance_2 + 
                         percent_dc +
                         percent_water + 
                         percent_herb + 
                         percent_canopy +
                         (1|hgm_2),
                       data = cattle_dc)
summary(m.dc.hoofpunch)


```




# 3: Comparing cattle impacts in early and late seasons
```{r timing}

# hoof punch
mtime.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ timing*closest_water_distance_2 + 
                              timing*percent_water + 
                              timing*percent_herb + 
                              timing*percent_canopy + 
                            (1|plot),
                            data = cattle_impact_repeat,
                            family = beta_family(link = "logit"))
summary(mtime.hoof.punch)

cattle_impact_repeat %>% 
  ggplot(aes(x = percent_water, y = percent_hoof_punch_2, 
             color = timing, fill = timing)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


# trampling
mtime.trample <- glmmTMB(percent_trample_2 ~ timing*closest_water_distance_2 + 
                              timing*percent_water + 
                              timing*percent_herb + 
                              timing*percent_canopy + 
                            (1|plot),
                            data = cattle_impact_repeat,
                            family = beta_family(link = "logit"))
summary(mtime.trample)


# manure
mtime.manure <- glmmTMB(manure_count ~ timing*closest_water_distance_2 + 
                              timing*percent_water + 
                              timing*percent_herb + 
                              timing*percent_canopy + 
                            (1|plot),
                            data = cattle_impact_repeat,
                            family = poisson(link = "log"))
summary(mtime.manure)

cattle_impact_repeat %>% 
  ggplot(aes(x = closest_water_distance_2, y = manure_count, 
             color = timing, fill = timing)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


# grazing intensity
mtime.grazing <- glmmTMB(grazing_high ~ timing*closest_water_distance_2 + 
                              timing*percent_water + 
                              timing*percent_herb + 
                              timing*percent_canopy + 
                            (1|plot),
                            data = cattle_impact_repeat,
                      family = binomial(link = "logit"))
summary(mtime.grazing)

cattle_impact_repeat %>% 
  ggplot(aes(x = closest_water_distance_2, y = grazing_high, 
             color = timing, fill = timing)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  theme_bw()

cattle_impact_repeat %>% 
  ggplot(aes(x = percent_water, y = grazing_high, 
             color = timing, fill = timing)) +
  geom_point(alpha=0.3) +
  geom_smooth(method = "lm") +
  theme_bw()



```


# 4: Cattle impacts on streams
What can we do with stream impact data? Would like to look at how cattle impact on nearby streams is impacted by restoration structures

Need to think about whether to include late survey data here too
```{r streams}

# filtering to plots where streams were present within 50m (stream impact data was recorded)
cattle_streams <- cattle_impact %>% 
  filter(!is.na(closest_water_distance)) %>%
  select(plot, hgm_2, watershed, timing,
         stream_impacts, 
         stream_incised:stream_trail,
         restoration_location, restoration_location_2, 
         stream_restoration_type,
         water_type_within_5m, water_type_within_10m)


length(unique(cattle_streams$plot))  # 132 plots with stream data recorded in early, 136 total for both visits

stream_summary <- cattle_streams %>% 
  group_by(timing, watershed) %>%
  summarise(
    n = n(),
    incised = sum(stream_incised),
    sloughing = sum(stream_sloughing),
    hoof_punch = sum(stream_hoof_punch),
    trampling = sum(stream_trample),
    manure = sum(stream_manure),
    grazing = sum(stream_grazing),
    trail = sum(stream_trail)
    )

stream_summary

# what else can we ask about cattle impacts on streams?

```


# 5: Subset plots in meadow areas with restoration structures for that analysis/comparison
Just remove the ones with nearby structures listed as "none"? Or subset based on geographical region?

```{r restoration-structures}

# next step -- analysis with restoration structures

stream_early <- cattle_streams %>% filter(timing == "Early")

# binomial model
m.incised <- glmmTMB(stream_incised ~ restoration_location + 
                      # stream_restoration_type +
                         watershed + (1|hgm_2),
                       data = stream_early, 
                       family = binomial(link = "logit"))
summary(m.incised)


m.sloughing <- glmmTMB(stream_sloughing ~ restoration_location + 
                      # stream_restoration_type +
                         watershed + (1|hgm_2),
                       data = stream_early, 
                       family = binomial(link = "logit"))
summary(m.sloughing)


m.hp <- glmmTMB(stream_hoof_punch ~ restoration_location + 
                      # stream_restoration_type +
                         watershed + (1|hgm_2),
                       data = stream_early, 
                       family = binomial(link = "logit"))
summary(m.hp)

```


# 6: Potential additional GIS components
- Distance to stream for plots with >max field distance
- Distribution up meadow vs down meadow for headwaters component



# Previous analyses and visualizations

```{r dist-water}


cattle_early %>%
  ggplot(aes(x = closest_water_distance_2, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


cattle_early %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()

cattle_late %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


# not sure how HGM categories fit into this model
m.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_early)
summary(m.hoof.punch)

# later plot sampling
# maybe want to do this as the subset of repeat plots with a repeated measures structure?
m.late.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_late)
summary(m.late.hoof.punch)

# trampling old linear model
m.trample <- lm(percent_trample ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_early)
summary(m.trample)






# Is the DC and percent water info overlapping?
# Correlated but actually not incredibly tight
cattle_early %>%
  ggplot(aes(x = percent_water, y = percent_dc)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()



```


Old DC and Hoof punch analyses

```{r dc-punch}

# Is hoof punch percent impact higher where DC is present?
cattle_early %>% 
  filter(!is.na(dc_present)) %>%
  ggplot(aes(x = dc_present, y = percent_hoof_punch, fill = dc_present)) +
  geom_boxplot() +
  theme_bw() +
  xlab("D. californica present") +
  ylab("Hoof Punch (% of plot)") +
  scale_fill_manual(values = c('goldenrod','darkgreen')) +
  guides(fill = "none")
ggsave("Figures/DC_Hoof-Punch_plot.pdf", width = 7.5, height = 5.5, units = "in")


# Framing 3a: How does hoof punch % correlate with DC % overall?
cattle_early %>% 
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


# Framing 3b: How does hoof punch % correlate with DC % where DC is present?
cattle_early %>% 
  filter(percent_dc > 0) %>%
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


```


Visualizations comparing early visit to late visit

*TODO: there's at least one row where the date is NA*
Could switch to the plots with line width to track shifts for the grazing intensity categorical one
```{r repeat-timing, echo=FALSE}

# comparing early to later

# DC Grazed
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_dc_grazed, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()
# more grazing on DC as time goes on

# Overall grazing intensity
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = grazing_intensity, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# % Trample
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_trample, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# % Hoof Punch
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_hoof_punch, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# Manure Count
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = manure_count, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# Indicators of how well plots are re-located:

# percent shrub
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_shrub, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# percent_rock
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_rock, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# canopy
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_canopy, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# percent_dc
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_dc, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


```