---
title: "CW_cattle-impacts_analysis"
output: html_document
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmmTMB)
library(lubridate)
library(ordinal)
library(patchwork)
library(emmeans)
```

## Reading in Data

To Do: decide how to consider streams beyond search distance from plot -- arbitrary large distance or use GIS to get spatial analysis distance
Or consider taking out the distance to water component, since so many plots have 0s with water in the plot

```{r data-cleaning}
cattle_impact_extra_gis <- read.csv("data/SRWC_cattle-impacts_capstone_extra-info_2026.csv") %>%
  rename(
    plot = Plot,
    watershed = Watershed,
    dist_downmeadow = NEAR_DIST,
    meadow_id = MEADOW_FID
  ) %>%
  select(plot, watershed, dist_downmeadow, meadow_id)

cattle_impact_full <- read.csv("data/SRWC_cattle-impacts_capstone_2026.csv") %>% # reading in csv
  rename_with(tolower) %>%
  rename(  # cleaning up and simplifying column names
    plot = plot._number,
    watershed = subwatrshed_location, 
    hgm = hydrogeomorphic_classification
  ) %>%
  mutate(  # Editing data in our columns for cleaning (use with caution!)
    watershed = str_trim(watershed), # remove whitespace
    grazing_intensity = str_trim(grazing_intensity),  
    # factoring grazing intensity
    grazing_intensity = factor(case_when(grazing_intensity == "Light" ~ "Low", 
                                         TRUE ~ grazing_intensity), 
                               levels = c("None", "Low", "Moderate", "High")), 
    closest_water_distance = as.numeric(closest_water_distance), # formatting Closest_Water_Distance column as a number
    date = as.Date(date, format = "%m/%d/%Y"), # Reformats the Date column into a format which R can interpret as dates correctly
    date = case_when(is.na(date) ~ as.Date("09/09/2025", format = "%m/%d/%Y"),
                     TRUE ~ date) # fixing three dates that came in with the wrong format
  ) %>%
  filter(!is.na(plot)) %>% # removing one blank row
  left_join(cattle_impact_extra_gis, by = join_by(plot, watershed)) # adding extra GIS info


cattle_impact <- cattle_impact_full  %>%
  #  select(date, plot, watershed, hgm, percent_canopy:manure_count, 
  #         closest_water_distance, water_type, restoration_location,
  #         stream_restoration_type:cattle_trail_crossing_or_bedding_area) %>%  # selecting relevant columns
  mutate(
    
    cattle_impact_on_stream = str_trim(cattle_impact_on_stream),
    
    hgm = str_trim(hgm),
    
    hgm_2 = case_when(
      hgm %in% c("Riparian_Mid_Gradient", "Riparian_High_Gradient") ~ "Riparian",
      hgm %in% c("Subsurface_Low_Gradient", "Subsurface_Mid_Gradient", "Subsurface_High_Gradient") ~ "Subsurface",
      TRUE ~ hgm),
    
    restoration_location = replace_na(str_trim(restoration_location), "None"),
    
    restoration_location_2 = case_when(
      restoration_location %in% c("Both", "Inside") ~ "in_or_both", 
      restoration_location == "Downtream" ~ "downstream",
      restoration_location == "Upstream" ~ "upstream",
      TRUE ~ "none"),
    
    stream_restoration_type = replace_na(str_trim(stream_restoration_type), "None"),
    
    hoof_punch_present = case_when(
      percent_hoof_punch > 0 ~ "yes",
      TRUE ~ "no"),
    
    percent_dc_2 = percent_dc/100, # matching proportions
    
    # proportions instead of percents for beta distribution models
    percent_hoof_punch_2 = 
      case_when(
        percent_hoof_punch == 0 ~ 0.001,
        TRUE ~ percent_hoof_punch/100),
    percent_trample_2 = 
      case_when(
        percent_trample == 0 ~ 0.001,
        TRUE ~ percent_trample/100),
    percent_dc_grazed_2 = case_when(
      percent_dc_grazed == 0 ~ 0.001,
      TRUE ~ percent_dc_grazed/100),
    
    percent_water_2 = percent_water/100,
    percent_herb_2 = percent_herb/100,
    percent_canopy_2 = percent_canopy/100,
    percent_shrub_2 = percent_shrub/100,
    percent_bare_2 = percent_bare/100,
    percent_litter_2 = percent_litter/100,
    
    # If stream distance wasn't recorded it was > 50m, so putting that in as our minimum for now
    closest_water_distance_2 = case_when(
      is.na(closest_water_distance) ~ 50, 
      TRUE ~ closest_water_distance
    ),
    
    # distance to streams specifically as a cattle water resource
    dist_stream = as.numeric( case_when(
      water_type_within_5m %in% c("Stream", "Both") ~ as.numeric(water_within_5m_dist),
      water_type_within_10m %in% c("Stream", "Both") ~ as.numeric(water_within_10m_or_more_dist),
      TRUE ~ 50
    )),
    
    # distance to stream as categorical
    dist_stream_3 = case_when(
      dist_stream <= 20 ~ "20m or less",
      dist_stream < 50 ~ "50m or less",
      TRUE ~ "greater than 50m"
    ),
    
    dist_stream_2 = case_when(
      dist_stream_3 == "20m or less" ~ "20m or less",
                              TRUE ~ "20m or more"
      ),
    
    # Binary version of grazing intensity
    grazing_high = case_when(
      grazing_intensity %in% c("None", "Low") ~ 0,
      grazing_intensity %in% c("Moderate", "High") ~ 1,
      TRUE ~ NA
    ),
    
    # splitting stream impacts into separate columns
    stream_impacts = tolower(stream_impacts),
    stream_impacts = replace_na(stream_impacts, ""), # some sites with no stream impacts imported as na rather than blank or none
    stream_incised = str_detect(stream_impacts, "inci"),
    stream_sloughing = str_detect(stream_impacts, "slough"),
    stream_grazing = str_detect(stream_impacts, "graz"),
    stream_trample = str_detect(stream_impacts, "tramp"),
    stream_hoof_punch = str_detect(stream_impacts, "hoof"),
    stream_manure = str_detect(stream_impacts, "manur"),
    stream_trail = str_detect(stream_impacts, "trail|cross"),
    
    
    # adding early vs late visit timing information
    timing = case_when(
      date > "2025-09-01" ~ "Late",
      date <= "2025-09-01" ~ "Early",
      TRUE ~ "NA"
    ),
    
    # switching distance downmeadow to km
    dist_downmeadow_km = dist_downmeadow/1000
  ) 


# early vist (all)
cattle_early <- cattle_impact %>%
  filter(timing == "Early")

# late visit (subset)
cattle_late <- cattle_impact %>%
  filter(timing == "Late")

# repeated plots both visits
late_plots <- cattle_impact %>%
  filter(timing == "Late") %>%
  pull(plot)

cattle_impact_repeat <- cattle_impact %>%
  filter(plot %in% late_plots) %>%
  arrange(plot) 


# details for plots
watershed.cols <- c("seagreen3", "darkorchid3")
watershed.labels = as_labeller(c(
  CM = "Cabin Meadows",
  RF = "Rock Fence"))



```

# 0: Summaries of cattle impacts

Want to also add information like percent water, canopy, water within 5m and within 10 m?
Would that go into the same table or a separate one?

Also do we want the late summaries here or are they just distracting?

What if the table design goes the other way, with columns under CM and RF for proportion of plots present, median, mean +- sd
and the rows are the different components of the data? 
```{r cattle-summaries}


plot_summary <- cattle_impact %>% 
  group_by(timing,watershed) %>%
  summarise(
    n = n(),
    across(starts_with(c("percent", "manure")), 
           list(present = ~ sum(.x > 0, na.rm = TRUE)/n,
                median = ~ median(.x, na.rm = TRUE), 
                mean = ~ mean(.x, na.rm = TRUE), 
                sd = ~ sd(.x, na.rm = TRUE))
    )
  ) 

plot_summary_long <- plot_summary %>%
  select(-n) %>%
  pivot_longer(
    cols = starts_with(c("percent", "manure"))
    ) %>%
  separate_wider_regex(name, c(component = ".*", "_", metric = ".*")) %>%
  pivot_wider(names_from = metric, values_from = value)


# formatting for table
plot_summary_meadow <- plot_summary_long %>% 
  pivot_wider(
    names_from = watershed, 
    values_from = present:sd,
    names_glue = "{watershed}_{.value}",
    names_vary = "slowest"
  ) %>%
  mutate(
    across(starts_with(c("CM", "RF")), ~ signif(.x, digits = 2)),
    CM_mean_sd = paste(CM_mean, "±", CM_sd),
    RF_mean_sd = paste(RF_mean, "±", RF_sd)
  ) %>%
  select(timing, component, CM_present, CM_median, CM_mean_sd,
         RF_present, RF_median, RF_mean_sd) 
  
# write plot composition table
plot_summary_meadow %>%
  filter(component %in% c("percent_canopy", "percent_herb",
                          "percent_shrub", "percent_bare", 
                          "percent_litter", "percent_rock",
                          "percent_woody","percent_water",
                          "percent_dc", "manure_count",
                          "percent_trample", "percent_hoof_punch",
                          "percent_dc_grazed")) %>%
  write.csv(file = "SRWC_cattle-impacts_plot-summary.csv")


# additional cattle components 
# (could pivot_longer this and rbind this in to cattle impact table)
cattle_additional <- cattle_impact %>% 
  group_by(timing, watershed) %>%
  summarise(
    n = n(),
    grazing_present = sum(grazing_intensity %in% c("Low", "Moderate", "High"))/n, # proportions of plots
    high_grazing_present = sum(grazing_intensity == "High")/n, 
  )

cattle_additional

```

Similar summaries but not separated by watershed or timing
```{r overall-summary}

overall_summary <- cattle_impact %>% 
#  group_by(timing,watershed) %>%
  summarise(
    n = n(),
    across(starts_with(c("percent", "manure")), list(present = ~ sum(.x > 0, na.rm = TRUE)/n,
                                        median = ~ median(.x, na.rm = TRUE), 
                                        mean = ~ mean(.x, na.rm = TRUE), 
                                        sd = ~ sd(.x, na.rm = TRUE),
                                        min = ~ min(.x, na.rm = TRUE),
                                        max = ~ max(.x, na.rm = TRUE))
    )
  ) 

overall_summary_long <- overall_summary %>%
  select(-n) %>%
  pivot_longer(
    cols = starts_with(c("percent", "manure"))
    ) %>%
  separate_wider_regex(name, c(component = ".*", "_", metric = ".*")) %>%
  pivot_wider(names_from = metric, values_from = value) 

# formatting for table
overall_summary_long %>% mutate(
    across(present:sd, ~ signif(.x, digits = 2)),
    mean_sd = paste(mean, "±", sd),
    range = paste(min, "—", max)
    ) %>% 
  select(component, present, median, mean_sd, range) %>%
  filter(component %in% c("percent_canopy", "percent_herb",
                          "percent_shrub", "percent_bare", 
                          "percent_litter", "percent_rock",
                          "percent_woody","percent_water",
                          "percent_dc", "manure_count",
                          "percent_trample", "percent_hoof_punch",
                          "percent_dc_grazed")) %>%
  write.csv(file = "SRWC_cattle-impacts_overall-summary.csv")

overall_cattle_additional <- cattle_impact %>% 
  summarise(
    n = n(),
    grazing_present = sum(grazing_intensity %in% c("Low", "Moderate", "High"))/n, # proportions of plots
    high_grazing_present = sum(grazing_intensity == "High")/n
  )

overall_cattle_additional


```



# 1: Predictors of cattle impacts
How to model grazing intensity? 

```{r cattle-impacts}


# hoof punch with beta distribution
glm.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ 
                          #  closest_water_distance_2 + 
                            dist_stream_2 +
                            percent_water_2 + 
                            percent_herb_2 + 
                            percent_canopy_2 + 
                            dist_downmeadow_km +
                            watershed +
                            (1|meadow_id),
                          data = cattle_early,
                          family = beta_family(link = "logit"))
summary(glm.hoof.punch)

hoof.punch.stream <- emmeans(glm.hoof.punch, specs = ~ dist_stream_2, type = "response")
hoof.punch.stream
contrast(hoof.punch.stream, method = "pairwise")


p.hoof.punch.canopy <- cattle_early %>% 
  ggplot(aes(x = percent_canopy, y = percent_hoof_punch, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Canopy cover (%)") +
  ylab("Hoof punch (%)") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none")

p.hoof.punch.herb <- cattle_early %>% 
  ggplot(aes(x = percent_herb, y = percent_hoof_punch, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Herbaceous cover (%)") +
  ylab("Hoof punch (%)") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none")

p.hoof.punch.water <- cattle_early %>% 
  ggplot(aes(x = percent_water, y = percent_hoof_punch, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Water cover (%)") +
  ylab("Hoof punch (%)") +
  geom_smooth(method = "lm", alpha = 0.2) +  
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25)) +
  theme_bw() +
  theme(legend.position = "none")



p.hoof.punch.water.2 <- cattle_early %>% 
  ggplot(aes(x = dist_stream_2, y = percent_hoof_punch, 
             color = watershed, fill = watershed)) +
  geom_violin(inherit.aes = FALSE, 
              aes(x = dist_stream_2, y = percent_hoof_punch),
              scale = "count",
              adjust = 2, linewidth = 0,
              fill = "grey50") +
  geom_point(position = position_jitter(width = 0.1, height = 0.1), alpha = 0.5) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Distance to stream (m)") +
  ylab("Hoof punch (%)") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.25, 0.8))

p.hoof.punch.canopy + p.hoof.punch.herb + p.hoof.punch.water + p.hoof.punch.water.2

ggsave("Figures/hoof-punch.pdf", width = 7, height = 6, units = "in")



# trampling with beta distribution
glm.trample <- glmmTMB(percent_trample_2 ~ 
                         dist_stream_2 +
                         percent_water_2 + 
                         percent_herb_2 + 
                         percent_canopy_2 + 
                         dist_downmeadow_km +
                         watershed +
                         (1|meadow_id),
                       data = cattle_early,
                       family = beta_family(link = "logit"))
summary(glm.trample)

trample.stream <- emmeans(glm.trample, specs = ~ dist_stream_2, type = "response")
trample.stream
contrast(trample.stream, method = "pairwise")


# manure with poisson distribution
glm.manure <- glmmTMB(manure_count ~ 
                        dist_stream_2 +
                        percent_water_2 + 
                        percent_herb_2 + 
                        percent_canopy_2 + 
                        dist_downmeadow_km +
                        watershed +
                        (1|meadow_id),
                      data = cattle_early, 
                      family = poisson(link = "log"))
summary(glm.manure)

manure.stream <- emmeans(glm.manure, specs = ~ dist_stream_2, type = "response")
manure.stream
contrast(manure.stream, method = "pairwise")

p.manure.canopy <- cattle_early %>% 
  ggplot(aes(x = percent_canopy, y = manure_count, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Canopy cover (%)") +
  ylab("Manure count") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none")

p.manure.herb <- cattle_early %>% 
  ggplot(aes(x = percent_herb, y = manure_count, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Herbaceous cover (%)") +
  ylab("Manure count") +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_x_continuous(breaks = c(0, 25, 50, 75, 100)) +
  theme_bw() +
  theme(legend.position = "none")

p.manure.water <- cattle_early %>% 
  ggplot(aes(x = percent_water, y = manure_count, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Water cover (%)") +
  ylab("Manure count") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none")

p.manure.water.2 <- cattle_early %>% 
  ggplot(aes(x = dist_stream_2, y = manure_count, 
             color = watershed, fill = watershed)) +
  geom_violin(inherit.aes = FALSE, 
              aes(x = dist_stream_2, y = manure_count),
              scale = "count",
              adjust = 2, linewidth = 0,
              fill = "grey50") +
  geom_point(position = position_jitter(width = 0.1, height = 0.1), alpha = 0.5) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Distance to stream (m)") +
  ylab("Manure count") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.25, 0.8))

p.manure.canopy + p.manure.herb + p.manure.water + p.manure.water.2
 ggsave("Figures/manure.pdf", width = 7, height = 6, units = "in")


# Ordinal grazing intensity data -- cumulative link models using the 'ordinal' package
# Not sure how to interpret this one yet
# could group to med-high and low-none if needed
m.grazing <- clmm(grazing_intensity ~ closest_water_distance_2 + 
                    percent_water + 
                    percent_herb + 
                    percent_canopy +
                    (1|hgm_2),
                  data = cattle_early) 
summary(m.grazing)

# none-low vs moderate-high grazing using a binomial model
m.grazing.2 <- glmmTMB(grazing_high ~ 
                         dist_stream_2 + 
                         percent_water + 
                         percent_herb + 
                         percent_canopy +
                         watershed +
                         (1|meadow_id),
                       data = cattle_early, 
                       family = binomial(link = "logit"))
summary(m.grazing.2)

grazing.stream <- emmeans(m.grazing.2, specs = ~ dist_stream_2, type = "response")
grazing.stream
contrast(grazing.stream, method = "pairwise")

```


# 2: Connecting cattle impacts to Darlingtonia
How do cattle impacts correlate with Darlingtonia presence (as a fen indicator)? Where are the biggest direct impacts on Darlingtonia (hoof punch and grazing of Darlingtonia)?
```{r dc}

# hoof punch is higher in sites with dc present (which are the wet sites)
dc.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ 
                           percent_dc_2 +
                           (1|meadow_id),
                         data = cattle_early,
                         family = beta_family(link = "logit"))
summary(dc.hoof.punch)


# trample doesn't differ by dc presence
dc.trample <- glmmTMB(percent_trample_2 ~ 
                        percent_dc_2 +
                        (1|meadow_id),
                      data = cattle_early,
                      family = beta_family(link = "logit"))
summary(dc.trample)

# manure is lower where dc is present
dc.manure <- glmmTMB(manure_count ~ 
                       percent_dc_2 +
                       (1|meadow_id),
                     data = cattle_early,
                     family = poisson(link = "log"))
summary(dc.manure)


# none-low vs moderate-high grazing using a binomial model
dc.grazing <- glmmTMB(grazing_high ~ 
                        percent_dc_2 +
                        (1|meadow_id),
                      data = cattle_early, 
                      family = binomial(link = "logit"))
summary(dc.grazing)

p.manure.dc <- cattle_early %>% 
  ggplot(aes(x = percent_dc, y = manure_count, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Darlingtonia cover (%)") +
  ylab("Manure count") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
 theme(legend.position = "none")

p.hoof.punch.dc <- cattle_early %>% 
  ggplot(aes(x = percent_dc, y = percent_hoof_punch, 
             color = watershed, fill = watershed)) +
  geom_point(position = position_jitter(width = 0.5, height = 0.5), alpha = 0.7) +
  scale_color_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  scale_fill_manual(values = watershed.cols, labels = watershed.labels, 
                     name = "Watershed") +
  xlab("Darlingtonia cover (%)") +
  ylab("Hoof Punch (%)") +
  geom_smooth(method = "lm", alpha = 0.2) +
  theme_bw() +
 theme(legend.position = "inside", legend.position.inside = c(0.75, 0.85))

p.manure.dc + p.hoof.punch.dc
ggsave("Figures/dc.pdf", width = 7, height = 4, units = "in")

```




# 3: Comparing cattle impacts in early and late seasons
Currently using all data collected at each time point -- ex percent water in the early season and (likely less) percent water in the late season. This is probably the best way to go about it, but it is worth noting that some of these components will change through time

For the repeat plots, probably want to just gorup dist to stream into two categories, since there are only a few in the 20m - 50m group
```{r timing}

timing_cols <- c('steelblue', 'darkgoldenrod')

# hoof punch
mtime.hoof.punch <- glmmTMB(percent_hoof_punch_2 ~ 
                              timing*dist_stream_2 + 
                              timing*percent_water +   
                              timing*percent_herb + 
                              timing*percent_canopy + 
                              watershed +
                            (1|meadow_id/plot),
                            data = cattle_impact_repeat,
                            family = beta_family(link = "logit"))
summary(mtime.hoof.punch)

p.time.hoof.punch <- cattle_impact_repeat %>% 
  ggplot(aes(x = percent_water, y = percent_hoof_punch_2, 
             color = timing, fill = timing)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = timing_cols) +
  scale_fill_manual(values = timing_cols) +
  xlab("Water cover (%)") +
  ylab("Hoof punch (%)") +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.8, 0.8))


# trampling
mtime.trample <- glmmTMB(percent_trample_2 ~ 
                           timing*dist_stream_2 + 
                           timing*percent_water + 
                           timing*percent_herb + 
                           timing*percent_canopy + 
                           (1|meadow_id/plot),
                         data = cattle_impact_repeat,
                         family = beta_family(link = "logit"))
summary(mtime.trample)


# manure
mtime.manure <- glmmTMB(manure_count ~ 
                          timing*dist_stream_2 + 
                          timing*percent_water + 
                          timing*percent_herb + 
                          timing*percent_canopy + 
                          (1|meadow_id/plot),
                        data = cattle_impact_repeat,
                        family = poisson(link = "log"))
summary(mtime.manure)

# comparisons of stream distance
manure.timing.emm <- emmeans(mtime.manure, 
                             specs = ~ dist_stream_2|timing, 
                             type = "response")

manure.timing.emm
contrast(manure.timing.emm, method = "pairwise")

p.time.manure <- cattle_impact_repeat %>% 
  ggplot(aes(x = dist_stream_2, y = manure_count, 
             fill = timing)) +
  geom_violin(scale = "count",
              adjust = 2, linewidth = 0,
              alpha = 0.7) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             jitter.height = 0.15,
                                             dodge.width = 0.9),
             alpha = 0.4) +
  scale_fill_manual(values = timing_cols) +
  xlab("Distance to stream (m)") +
  ylab("Manure count") +
  theme_bw() +
  theme(legend.position = "none")

p.time.manure.water <- cattle_impact_repeat %>% 
  ggplot(aes(x = percent_water, y = manure_count, 
             color = timing, fill = timing)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = timing_cols) +
  scale_fill_manual(values = timing_cols) +
  xlab("Water cover (%)") +
  ylab("Manure count") +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25)) +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.8, 0.8))


# grazing intensity
mtime.grazing <- glmmTMB(grazing_high ~ 
                           timing*dist_stream_2 + 
                           timing*percent_water + 
                           timing*percent_herb + 
                           timing*percent_canopy + 
                           (1|meadow_id/plot),
                         data = cattle_impact_repeat,
                         family = binomial(link = "logit"))
summary(mtime.grazing)

# comparisons of stream distance
grazing.timing.emm <- emmeans(mtime.grazing, 
                             specs = ~ dist_stream_2|timing)

grazing.timing.emm
contrast(grazing.timing.emm, method = "pairwise")

p.time.grazing.herb <- cattle_impact_repeat %>% 
  ggplot(aes(x = percent_herb, y = grazing_high, 
             color = timing, fill = timing)) +
  geom_point(position = position_dodge(width = 3),
             alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = timing_cols) +
  scale_fill_manual(values = timing_cols) +
  xlab("Herbaceous cover (%)") +
  ylab("High grazing intensity") +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 25, 50, 75, 100)) +
  theme(legend.position = "none")

p.time.grazing.dist <- cattle_impact_repeat %>% 
  ggplot(aes(x = dist_stream_2, y = grazing_high, 
             fill = timing)) +
    geom_violin(scale = "count",
              adjust = 2, linewidth = 0,
              alpha = 0.7) +
  geom_point(position = position_jitterdodge(jitter.width = 0.3,
                                             jitter.height = 0.05,
                                             dodge.width = 0.9),
             alpha = 0.5) +
  scale_fill_manual(values = timing_cols) +
  scale_color_manual(values = timing_cols) +
  xlab("Distance to stream (m)") +
  ylab("High grazing intensity") +
  theme_bw() +
  theme(legend.position = "none")

p.time.manure.water + p.time.manure 
ggsave("Figures/timing.pdf", width = 7, height = 4, units = "in")


```


# 4: Cattle impacts on streams
What can we do with stream impact data? Would like to look at how cattle impact on nearby streams is impacted by restoration structures

Need to think about whether to include late survey data here too
```{r streams}

# filtering to plots where streams were present within 50m (stream impact data was recorded)
cattle_streams <- cattle_impact %>% 
  filter(!is.na(closest_water_distance)) %>%
  select(plot, meadow_id, hgm_2, watershed, timing,
         stream_impacts, 
         stream_incised:stream_trail,
         restoration_location, restoration_location_2, 
         stream_restoration_type,
         water_type_within_5m, water_type_within_10m,
         dist_downmeadow_km) %>%
  mutate(stream_impact_any = stream_incised | stream_sloughing | stream_grazing | stream_trample | stream_hoof_punch | stream_manure | stream_trail)


cattle_streams %>% 
  filter(stream_impact_any == TRUE) %>%
  pull(plot) %>%
  unique() %>%
  length()
# 117 plots with stream data recorded in early, 121 total for both visits

stream_summary <- cattle_streams %>% 
  group_by(timing, watershed) %>%
  summarise(
    n = n(),
    any = sum(stream_impact_any)/n,
    incised = sum(stream_incised)/n,
    sloughing = sum(stream_sloughing)/n,
    hoof_punch = sum(stream_hoof_punch)/n,
    trampling = sum(stream_trample)/n,
    manure = sum(stream_manure)/n,
    grazing = sum(stream_grazing)/n,
    trail = sum(stream_trail)/n
    )

stream_summary_long <- stream_summary %>%
  select(-n) %>%
  pivot_longer(
    cols = any:trail
  ) %>%
  pivot_wider(
    names_from = watershed, 
    values_from = value,
  ) %>%
  mutate(
    across(starts_with(c("CM", "RF")), ~ signif(.x, digits = 2)),
  )
# write.csv(stream_summary_long, file = "SRWC_cattle-impacts_stream-summary.csv") 



```


# 5: Subset plots in meadow areas with restoration structures for that analysis/comparison
Just remove the ones with nearby structures listed as "none"? Or subset based on geographical region?

```{r restoration-structures}

# next step -- analysis with restoration structures

stream_early <- cattle_streams %>% filter(timing == "Early")

# binomial model
m.incised <- glmmTMB(stream_incised ~ restoration_location_2 + 
                       dist_downmeadow_km +
                       watershed + (1|meadow_id),
                     data = stream_early, 
                     family = binomial(link = "logit"))
summary(m.incised)

incised.emm <- emmeans(m.incised, specs = ~ restoration_location_2, type = "response")
incised.emm
contrast(incised.emm, method = "pairwise")


m.sloughing <- glmmTMB(stream_sloughing ~ restoration_location_2 + 
                      dist_downmeadow_km +
                         watershed + (1|meadow_id),
                       data = stream_early, 
                       family = binomial(link = "logit"))
summary(m.sloughing)
slough.emm <- emmeans(m.sloughing, specs = ~ restoration_location_2, type = "response")
slough.emm
contrast(slough.emm, method = "pairwise")


m.hp <- glmmTMB(stream_hoof_punch ~ restoration_location_2 + 
                  dist_downmeadow_km +
                  watershed + (1|meadow_id),
                data = stream_early, 
                family = binomial(link = "logit"))
summary(m.hp)

m.t <- glmmTMB(stream_trample ~ restoration_location_2 + 
                  dist_downmeadow_km +
                  watershed + (1|meadow_id),
                data = stream_early, 
                family = binomial(link = "logit"))
summary(m.t)

# could look at a combination of any impacts (incised, sloughing, or hoof punch)

```




# Previous analyses and visualizations

```{r dist-water}


cattle_early %>%
  ggplot(aes(x = closest_water_distance_2, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


cattle_early %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()

cattle_late %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


# not sure how HGM categories fit into this model
m.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_early)
summary(m.hoof.punch)

# later plot sampling
# maybe want to do this as the subset of repeat plots with a repeated measures structure?
m.late.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_late)
summary(m.late.hoof.punch)

# trampling old linear model
m.trample <- lm(percent_trample ~ closest_water_distance_2 + percent_water + 
                        percent_herb + percent_canopy + percent_dc,
                    data = cattle_early)
summary(m.trample)






# Is the DC and percent water info overlapping?
# Correlated but actually not incredibly tight
cattle_early %>%
  ggplot(aes(x = percent_water, y = percent_dc)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()



```


Old DC and Hoof punch analyses

```{r dc-punch}

# Is hoof punch percent impact higher where DC is present?
cattle_early %>% 
  filter(!is.na(dc_present)) %>%
  ggplot(aes(x = dc_present, y = percent_hoof_punch, fill = dc_present)) +
  geom_boxplot() +
  theme_bw() +
  xlab("D. californica present") +
  ylab("Hoof Punch (% of plot)") +
  scale_fill_manual(values = c('goldenrod','darkgreen')) +
  guides(fill = "none")
ggsave("Figures/DC_Hoof-Punch_plot.pdf", width = 7.5, height = 5.5, units = "in")


# Framing 3a: How does hoof punch % correlate with DC % overall?
cattle_early %>% 
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


# Framing 3b: How does hoof punch % correlate with DC % where DC is present?
cattle_early %>% 
  filter(percent_dc > 0) %>%
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


```


Visualizations comparing early visit to late visit

*TODO: there's at least one row where the date is NA*
Could switch to the plots with line width to track shifts for the grazing intensity categorical one
```{r repeat-timing, echo=FALSE}

# comparing early to later

# DC Grazed
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_dc_grazed, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()
# more grazing on DC as time goes on

# Overall grazing intensity
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = grazing_intensity, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# % Trample
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_trample, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# % Hoof Punch
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_hoof_punch, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# Manure Count
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = manure_count, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


# Indicators of how well plots are re-located:

# percent shrub
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_shrub, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# percent_rock
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_rock, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# canopy
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_canopy, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()

# percent_dc
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_dc, group = plot, color = hgm_2)) +
  geom_line(position = position_jitter()) +
  theme_bw()


```