---
title: "CW_cattle-impacts_analysis"
output: html_document
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Reading in Data

To Do: decide how to consider streams beyond search distance from plot -- arbitrary large distance or use GIS to get spatial analysis distance

```{r cars}
cattle_impact_full <- read.csv("Cleaned_Capstone_Data.csv") %>% # reading in csv
  rename_with(tolower) %>%
  rename(  # cleaning up and simplifying column names
    plot = plot._number,
    watershed = subwatrshed_location, 
    hgm = hydrogeomorphic_classification
  ) %>%
  mutate(  # Editing data in our columns for cleaning (use with caution!)
    watershed = str_trim(watershed), # remove whitespace from Watershed column
    closest_water_distance = as.numeric(closest_water_distance), # formatting Closest_Water_Distance column as a number
    date = as.Date(date, format = "%m/%d/%Y") # Reformats the Date column into a format which R can interpret as dates correctly
  ) %>%
  filter(!is.na(plot)) # removing one blank row


cattle_impact <- cattle_impact_full  %>%
  select(date, plot, watershed, hgm, percent_canopy:manure_count, 
         closest_water_distance, water_type, restoration_location,
         stream_restoration_type:cattle_trail_crossing_or_bedding_area) %>%  # selecting relevant columns
  mutate(
    restoration_location_2 = case_when(
      restoration_location %in% c("Both", "Inside", "Downstream") ~ "in_or_downstream", 
      TRUE ~ "none_or_upstream"),
    
    cattle_impact_on_stream = str_trim(cattle_impact_on_stream),
    
    hgm = str_trim(hgm),
    
    hgm_2 = case_when(
      hgm %in% c("Riparian_Mid_Gradient", "Riparian_High_Gradient") ~ "Riparian",
      hgm %in% c("Subsurface_Low_Gradient", "Subsurface_Mid_Gradient", "Subsurface_High_Gradient") ~ "Subsurface",
      TRUE ~ hgm),
    
    hoof_punch_present = case_when(
      percent_hoof_punch > 0 ~ "yes",
      TRUE ~ "no"),
    
    # If stream distance wasn't recorded it was > 50m, so putting that in as our minimum for now
    closest_water_distance_2 = case_when(
      is.na(closest_water_distance) ~ 50, 
      TRUE ~ closest_water_distance
    )
  )


# early vist (all)
cattle_early <- cattle_impact %>%
  filter(date < "2025-09-01")

# late visit (subset)
cattle_late <- cattle_impact %>%
  filter(date > "2025-09-01")

# repeated plots both visits
late_plots <- cattle_impact_full %>%
  filter(date > "2025-09-01") %>%
  pull(plot)

cattle_impact_repeat <- cattle_impact %>%
  filter(plot %in% late_plots) %>%
  mutate(
    timing = case_when(
      date > "2025-09-01" ~ "Late",
      date <= "2025-09-01" ~ "Early",
      TRUE ~ "NA"
    )
  ) %>%
arrange(plot)


```

## Comparing early visit to late visit

*TODO: there's at least one row where the date is NA*

```{r repeat-timing, echo=FALSE}

# comparing early to later

# % DC
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_dc, group = plot, color = hgm_2)) +
  geom_line() +
  theme_bw()

# % Trample
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_trample, group = plot, color = hgm_2)) +
  geom_line() +
  theme_bw()


# % Hoof Punch
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = percent_hoof_punch, group = plot, color = hgm_2)) +
  geom_line() +
  theme_bw()


# Manure Count
cattle_impact_repeat %>%
  ggplot(aes(x = timing, y = manure_count, group = plot, color = hgm_2)) +
  geom_line() +
  theme_bw()



```


CW thoughts/notes of interest -- percent canopy or percent herb vs cattle impacts?


# DC and Hoof punch analyses

```{r dc-punch}

# Is hoof punch percent impact higher where DC is present?
cattle_early %>% 
  mutate(dc_present = percent_DC > 0) %>%
  filter(!is.na(dc_present)) %>%
  ggplot(aes(x = dc_present, y = percent_hoof_punch, fill = dc_present)) +
  geom_boxplot() +
  theme_bw() +
  xlab("D. californica present") +
  ylab("Hoof Punch (% of plot)") +
  scale_fill_manual(values = c('goldenrod','darkgreen')) +
  guides(fill = "none")
ggsave("Figures/DC_Hoof-Punch_plot.pdf", width = 7.5, height = 5.5, units = "in")


# Framing 3a: How does hoof punch % correlate with DC % overall?
cattle_early %>% 
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


# Framing 3b: How does hoof punch % correlate with DC % where DC is present?
cattle_early %>% 
  filter(percent_dc > 0) %>%
  ggplot(aes(x = percent_dc, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm")


```


# Broader analyses understanding cattle impacts and water

Distance to stream
Type of water/stream width?
Location within the meadow (south/headwaters)?

How to model grazing intensity? 


What can we do with stream impact data? Would like to look at how cattle impact on nearby streams is impacted by restoration structures


Switch to beta distributions for percent cover variables?
```{r dist-water}


cattle_early %>%
  ggplot(aes(x = closest_water_distance_2, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


cattle_early %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()

cattle_late %>%
  ggplot(aes(x = percent_water, y = percent_hoof_punch)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


# not sure how HGM categories fit into this model
m.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                       restoration_location_2,
                    data = cattle_early)
summary(m.hoof.punch)

# later plot sampling
m.late.hoof.punch <- lm(percent_hoof_punch ~ closest_water_distance_2 + percent_water + 
                       restoration_location_2,
                    data = cattle_late)
summary(m.late.hoof.punch)



# Is the DC and percent water info overlapping?
# Correlated but actually not incredibly tight
Cattle_Early %>%
  ggplot(aes(x = percent_water, y = percent_dc)) +
  geom_point(position = position_jitter()) +
  geom_smooth(method = "lm") +
  theme_bw()


m.trample <- lm(percent_trample ~ closest_water_distance_2 + percent_water + 
                       restoration_location_2,
                    data = cattle_early)
summary(m.trample)


m.manure <- lm(manure_count ~ closest_water_distance_2 + percent_water + 
                       restoration_location_2,
                    data = cattle_early)
summary(m.manure)



```

