---
title: "SRWC-seedbank_greenhouse_prelim-analysis"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)

# Color Options:

color.palatte <- c("mediumblue", "darkred") 
fun.palatte <- c("darkolivegreen","darkgoldenrod")
```

## Read in Data

#csv file for SRWC seedbank greenhouse study

```{r read-files}

##greenhouse.n <- SRWC_seedbank_greenhouse_2024_02_21_n
  
greenhouse.n <- read.csv(here("data/SRWC-seedbank_greenhouse_2024-05-22_n.csv"), header = TRUE)

greenhouse.n$X <- NULL


```



## Summarize for species richness

summarize by date
```{r summarize-date}

# switch to long format (same data)
greenhouse.long <- greenhouse.n %>% 
  pivot_longer(cols = rf.1.a:control, names_to = "plot", values_to = "count")

# summary by date
greenhouse.summary <- greenhouse.long %>% 
  group_by(date, observer, plot) %>% 
  dplyr::summarise(
    Richness = sum(count>0, na.rm = TRUE))

View(greenhouse.summary)

```
summarize across dates for max species richness
```{r summarize-date}

# switch to long format (same data)
greenhouse.long <- greenhouse.n %>% 
  pivot_longer(cols = rf.1.a:control, names_to = "plot", values_to = "count")

# summary by species max count
greenhouse.sum <- greenhouse.long %>% 
  group_by(species, plot) %>% 
  dplyr::summarise(
    count.max = max(count, na.rm = TRUE)) %>%
  filter(count.max > 0)

#View(greenhouse.sum)

# remove species present in control trays
control.species <- greenhouse.sum %>% 
  filter(plot == "control") %>% 
  select(species) %>% 
  unlist()

greenhouse.sum.clean <- greenhouse.sum %>%
  filter(!(species %in% control.species))


# summarize by plots for richness
greenhouse.max.summary <- greenhouse.sum.clean %>%
  group_by(plot) %>%
  dplyr::summarise(
    greenhouse.richness = sum(count.max > 0, na.rm = TRUE)
  )

greenhouse.max.summary


```

# Field data
Read it in, organize and summarize it in the same way to match the greenhouse.max.summary dataset



# Combine field and greenhouse data
Use the function left_join(dataframe1, dataframe2, by = "plot")
```{r read-data-2}

##field.data <- SRWC_seedbank_plant_survey_2024_02_26

field.data.raw <- read.csv(here("data/SRWC-seedbank_plant-survey_2024-02-26.csv"), header = TRUE)

# removing duplicate rows
dup.1 <- which(field.data.raw$plot == "rf.6.a" & field.data.raw$species.code == "unkfb")[2]
dup.2 <- which(field.data.raw$plot == "rf.4.p" & field.data.raw$species.code == "sidore")[2]
drop <- c(dup.1, dup.2)

field.data <- field.data.raw %>% filter(!row_number() %in% drop)


```

Summarize field data 

```{r summarize-data-2}


field.summary <- field.data %>%
  group_by(plot, habitat) %>%
  dplyr::summarise(
    field.richness = sum(species.count > 0, na.rm = TRUE))

field.summary

```

# Combine field and greenhouse data
Use the function left_join(dataframe1, dataframe2, by = "plot")
```{r join-data-1}

combined.sum <- left_join(greenhouse.max.summary, field.summary, by = "plot")

```

# Omit missing data (control)
```{r omit-data-1}

combined.summary <- na.omit(combined.sum)

```


# Begin linear regression:
```{r regress-data-1}

combined.lm = lm(greenhouse.richness ~ field.richness, data = combined.summary)

combined.lm

summary(combined.lm)
```
# Begin linear regression:
```{r graph-regression-1}

GH.vs.field.regression <- ggplot(combined.summary, aes(x=field.richness, y=greenhouse.richness)) +
    geom_point()+
    theme_classic()+
    xlab("Field Richness")+
    ylab("Greehouse Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

GH.vs.field.regression

ggsave(plot = GH.vs.field.regression, filename = "figures/GH-vs-field-regression.png", width = 6, height = 4, units = "in")

```
# Do an anova to compare habitat type:

# Field - Looking at distribution:
```{r anova-1}

boxplot(field.richness~habitat, data=combined.summary, main="Field Richness by Habitat Type", xlab = "Habitat", ylab = "Species Richness")
```
```{r ggboxplot-1}
  
field.richness.by.habitat <- combined.summary %>%
      ggplot(aes(x=habitat, y=field.richness))+
      geom_boxplot(aes(fill=habitat, col = fun.palatte), colour="black") +
      labs(x= "Habitat Type", y="Field Species Richness")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow", "N/A")) +
        scale_y_continuous(limits=c(5,25), breaks=c(5, 10, 15, 20, 25))+
        scale_fill_manual(values= fun.palatte) +
  theme(legend.position = "none")

field.richness.by.habitat

ggsave(plot = field.richness.by.habitat, filename = "figures/field_richness_by_habitat.png", width = 6, height = 4, units = "in")
```

#Field ANOVA time:
```{r anova-4}

fieldrichness.aov = aov(field.richness~habitat, data=combined.summary)

  summary(fieldrichness.aov)
  
```
# Greenhouse - Looking at distribution:
```{r anova-5}

boxplot(greenhouse.richness~habitat, data=combined.summary, main="Greenhouse Richness by Habitat Type", xlab = "Habitat", ylab = "Species Richness") 
 
```
#Greehouse ANOVA time:
```{r ggboxplot-2}
  
GH.richness.by.habitat <- combined.summary %>%
      ggplot(aes(x=habitat, y=greenhouse.richness))+
      geom_boxplot(aes(fill=habitat, col = color.palatte), colour="black") +
      labs(x= "Habitat Type", y="Greenhouse Species Richness")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow", "N/A")) +
        scale_y_continuous(limits=c(0,15), breaks=c(0, 5, 10, 15))+
        scale_fill_manual(values= fun.palatte)+
        theme(legend.position = "none")

GH.richness.by.habitat

ggsave(plot = GH.richness.by.habitat, filename = "figures/GH_richness_by_habitat.png", width = 6, height = 4, units = "in")

```

#Greehouse ANOVA time:

```{r anova-6}

greenhouserichness.aov = aov(greenhouse.richness~habitat, data=combined.summary)

  summary(greenhouserichness.aov)
  
```
#Two way ANOVA?:
```{r anova-7}

habitat.richness = aov(field.richness ~ habitat*greenhouse.richness, data=combined.summary)

summary(habitat.richness)
  
```
#Soil moisture 
```{r sm-read data}

##soil.moisture <- SRWC_seedbank_soil_moisture_2023_09_20

soil.moisture <- read.csv(here("data/SRWC-seedbank_soil-moisture_2023-09-20.csv"), header = TRUE)

combined.summary <- left_join(combined.summary, soil.moisture, by = c("plot","habitat"))
  
```

#soil moisture summary 

```{r SM-summarize-data}

moisture.lm = lm(moisture.content ~ habitat, data = soil.moisture)

moisture.lm

summary(moisture.lm)

```
#soil moisture ANOVA boxplot 
```{r SM-ggboxplot-1}
  
soil.moisture.plot <- soil.moisture %>%
      ggplot(aes(x=habitat, y=moisture.content))+
      geom_boxplot(aes(fill=habitat, col = color.palatte), colour="black") +
      labs(x= "Habitat Type", y="Soil Moisture Content (%)")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow")) +
        scale_y_continuous(limits=c(0,100), breaks=c(0, 25, 50, 75, 100))+
        scale_fill_manual(values= fun.palatte)+
        theme(legend.position = "none")

soil.moisture.plot

ggsave(plot = soil.moisture.plot, filename = "figures/soil-moisture-plot.png", width = 6, height = 4, units = "in")
    
```

#soil moisture ANOVA time!! 

```{r anova-8}

soilmoisture.aov = aov(moisture.content~habitat, data=soil.moisture)

  summary(soilmoisture.aov)
  
```
```{r SM-regress-data}

SMfield.lm = lm(field.richness ~ moisture.content, data=combined.summary)

SMfield.lm

summary(SMfield.lm)

```
```{r SM-regress-data-2}

SMGH.lm = lm(greenhouse.richness ~ moisture.content, data=combined.summary)

SMGH.lm

summary(SMGH.lm)

```
```{r anova-9}

habitat.richness = aov(field.richness ~ habitat*moisture.content, data=combined.summary)

summary(habitat.richness)

habitat.richness.2 = aov(greenhouse.richness ~ habitat*moisture.content, data=combined.summary)

summary(habitat.richness.2)
  
```

```{r anova-10}

habitat.names <- c(a="Actual Meadow", p="Potential Meadow")

field.richness.vs.soil.moisture <- ggplot(combined.summary, aes(x=moisture.content, y=field.richness)) +
  facet_wrap(vars(habitat),labeller=labeller(habitat=habitat.names), scales="free_x")+
    geom_point()+
    theme_classic()+
    xlab("Soil Moisture Content (%)")+
    ylab("Field Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

field.richness.vs.soil.moisture

ggsave(plot = field.richness.vs.soil.moisture, filename = "figures/field_richness_vs_soil_moisture.png", width = 6, height = 4, units = "in")

GH.richness.vs.soil.moisture <- ggplot(combined.summary, aes(x=moisture.content, y=greenhouse.richness)) +
  facet_wrap(vars(habitat),labeller=labeller(habitat=habitat.names), scales="free_x")+
    geom_point()+
    theme_classic()+
    xlab("Soil Moisture Content (%)")+
    ylab("Greenhouse Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

GH.richness.vs.soil.moisture

ggsave(plot = GH.richness.vs.soil.moisture, filename = "figures/GH_richness_vs_soil_moisture.png", width = 6, height = 4, units = "in")


```