---
title: "SRWC-cattle-exclosures_analysis"
output: html_document
date: "2026-05-22"
---

*TO DO* Add in hoof punch? Measured separately from trampling in 2026, could combine for consistency through time and/or examine it separately at the end point

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(glmmTMB)
library(emmeans)
library(patchwork)

# list of wet meadow vs dry meadow blocks
wet.blocks <- c(2, 3, 5, 6, 8)

cattle.ex <- read.csv("data/SRWC-cattle-exclosure_post-treatment_2026-09-04.csv") %>%
  rename_with(tolower) %>%
  rename(manure.count = x..cowpies, 
         grazed.perc = x..grazed, 
         height.1 = height.1..cm.) %>%
  separate_wider_delim(plot, " ", names = c("position", "treatment")) %>%
  mutate(date = as.Date(date, "%m/%d/%y"),
         plot = paste0(block, position),
         height.mean = (height.1 + height.2 + height.3 + height.4)/4,
         meadow.type = case_when(
           block %in% wet.blocks ~ "wet", 
           TRUE ~ "dry"),
         trample.both = case_when(  # split hoof punch and trampling in 2026, not in previous years
           is.na(hoof_punch) ~ trample,
           TRUE ~ trample + hoof_punch
         ),
         year = year(date),
         timing = case_when(
           month(date) == 9 ~ "end of season",
           month(date) == 8 & day(date) == 31 ~ "end of season",
           TRUE ~ "mid season"
         ),
         watershed = case_when(
           block %in% 1:5 ~ "CM", 
           block %in% 6:9 ~ "RF",
           TRUE ~ NA
         )
  )



# compare versus control in each plot at each time step for height, veg.cover, 
# animal.trample, water, darlingtonia, grazed.perc
cattle.control <- cattle.ex %>%
  group_by(block, date) %>%
  mutate(
    height.control = height.mean[treatment == "control"][1],
    height.diff = height.mean - height.control,
    
    veg.control = veg.cover[treatment == "control"][1],
    veg.diff = veg.cover - veg.control,
    
    litter.control = litter[treatment == "control"][1],
    litter.diff = litter - litter.control,
    
    trample.control = trample.both[treatment == "control"][1],
    trample.diff = trample.both - trample.control,
    
    water.control = water[treatment == "control"][1],
    water.diff = water - water.control,
    
    dc.control = darlingtonia[treatment == "control"][1],
    dc.diff = darlingtonia - dc.control,
    
    grazed.control = grazed.perc[treatment == "control"][1],
    grazed.diff = grazed.perc - grazed.control,
    
    manure.control = manure.count[treatment == "control"][1],
    manure.diff = manure.count - manure.control
  ) %>%
  ungroup()

# Compare to first year data, September visit only for now
# add water for this too?
cattle.change <- cattle.ex %>% 
  filter(timing == "end of season") %>%
  group_by(plot) %>%
  mutate(
    first.date = min(date),
    dc.start = darlingtonia[date == first.date][1],
    dc.change = darlingtonia - dc.start
  )

  
```

## Scattershot visualizations
Currently for each block, comparing to the control using just a difference -- could also think about a proportion response or a log response ratio

Or we can try to compare first year's data to third year's data and look at what changed, but it's also interesting to see the difference in ex trampling effects. Also need to think about how we're managing the two time points of data each year


```{r visualizations-control}

treat.cols.3 <- c("turquoise3","darkblue","dodgerblue")
treat.cols.4 <- c("salmon2", "turquoise3","darkblue","dodgerblue")

# trampling impact compared to control
p.trample <- cattle.control %>% 
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = trample.diff, 
             color = treatment, shape = meadow.type, 
             group = plot)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 0.5),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Trample % cover difference from control plot") +
  theme_bw() +
  theme(legend.box = "horizontal", legend.position = c(0.4, 0.85),
        legend.background = element_rect(fill = "transparent"),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9.5),
        legend.key.size = unit(0.4, "cm")) 


# grazed percent compared to control
p.grazed <- cattle.control %>%
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = grazed.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 0.5),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Grazing % cover difference from control plot") +
  theme_bw() +
  theme(legend.position = "none")


# veg height compared to control
p.veg.height <- cattle.control %>%
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = height.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 0.5),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Vegetation height difference from control plot") +
  theme_bw() +
  theme(legend.position = "none")


# veg cover compared to control
p.veg.cover <- cattle.control %>% 
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = veg.diff, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 1),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  #  geom_line(position = position_dodge(width = 50)) +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Vegetation % cover difference from control plot") +
  theme_bw() +
  theme(legend.position = "none")

# litter cover compared to control
p.litter <- cattle.control %>% 
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = litter.diff, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 1),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  #  geom_line(position = position_dodge(width = 50)) +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Litter % cover difference from control plot") +
  theme_bw() +
  theme(legend.position = "none")


# manure count compared to control
p.manure <- cattle.control %>% 
  filter(treatment %in% c("early", "late", "full")) %>%
  ggplot(aes(x = date, y = manure.diff, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_jitterdodge(dodge.width = 50, jitter.width = 10, jitter.height = 0.1),
             size = 2, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  #  geom_line(position = position_dodge(width = 50)) +
  scale_color_manual(values = treat.cols.3, name = "Exclosure treatment") +
  scale_shape_manual(values = c(16, 17), name = "Meadow type") +
  xlab("Survey Date") +
  ylab("Manure count difference from control plot") +
  theme_bw() +
  theme(legend.position = c("none"))

# putting plots together
p.manure + p.trample + p.grazed + p.veg.height + p.veg.cover + p.litter +
  plot_layout(nrow = 3)
ggsave(filename = "Figures/cattle-exclosures.pdf", width = 8, height = 10)




# Dc cover raw data through time
p.dc <- cattle.control %>% 
  filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = darlingtonia, 
             color = treatment, shape = meadow.type, group = plot)) +
#  geom_point(position = position_dodge(width = 30)) +
  geom_line(position = position_jitter(width = 3, height = 0.5)) +
  theme_bw() +
  scale_color_manual(values = treat.cols.4) +
  xlab("Survey Date") +
  ylab("D. californica cover") +
  theme(legend.position = "none")

# water cover raw data through time
p.water <- cattle.control %>% 
  filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = water, 
             color = treatment, shape = meadow.type, group = plot)) +
 # geom_point(position = position_jitter(width = 3, height = 1)) +
  geom_line(position = position_jitter(width = 3, height = 0.5)) +
  theme_bw() +
  scale_color_manual(values = treat.cols.4)  +
  xlab("Survey Date") +
  ylab("Water % cover") 

p.water / p.dc
# ggsave(filename = "Figures/cattle-exclosures_water-dc.pdf", width = 6, height = 5.5)

# Dc cover change through time
cattle.change %>%
  filter(meadow.type == "wet") %>%
  ggplot(aes(x = date, y = dc.change, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  geom_line(position = position_dodge(width = 30)) +
  theme_bw() +
  scale_color_manual(values = treat.cols.4)




```


# Analyses through time
Also switching to a beta distribution for the models, which is probably a good idea even for the single time-point for the cover-based metrics
```{r time-series}
cattle.control <- cattle.control %>%
  mutate(
    trample.2 = case_when(
      trample.both == 0 ~ 0.001,
      trample.both == 100 ~ 0.999,
      TRUE ~ trample.both/100),
    
    darlingtonia.2 = case_when(
      darlingtonia == 0 ~ 0.001,
      darlingtonia == 100 ~ 0.999,
      TRUE ~ darlingtonia/100),
    
    veg.2 = case_when(
      veg.cover == 0 ~ 0.001,
      veg.cover == 100 ~ 0.999,
      TRUE ~ veg.cover/100),
    
    litter.2 = case_when(
      litter == 0 ~ 0.001,
      litter  == 100 ~ 0.999,
      TRUE ~ litter /100),
    
    water.2 = case_when(
      water == 0 ~ 0.001,
      water == 100 ~ 0.999,
      TRUE ~ water /100),
    
    grazed.2 = case_when(
      grazed.perc== 0 ~ 0.001,
      grazed.perc == 100 ~ 0.999,
      TRUE ~ grazed.perc /100)
    
  )


# grazing across meadow types and years
m.time.grazed <- glmmTMB(grazed.2 ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.grazed)
emm.time.grazed <- emmeans(m.time.grazed, specs = ~ treatment,
                            type = "response")
emm.time.grazed
contrast(emm.time.grazed, method = "pairwise")

# trampling across meadow types and years
m.time.trample <- glmmTMB(trample.2 ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.trample)
emm.time.trample <- emmeans(m.time.trample, specs = ~ treatment,
                            type = "response")
emm.time.trample
contrast(emm.time.trample, method = "pairwise")

# trampling split by meadow type
m.time.trample.meadow <- glmmTMB(trample.2 ~ treatment*meadow.type + (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.trample.meadow) # interaction not significant

# veg height
m.time.veg <- glmmTMB(height.mean ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = cattle.control)
summary(m.time.veg)
emm.time.veg <- emmeans(m.time.veg, specs = ~ treatment,
                            type = "response")
emm.time.veg
contrast(emm.time.veg, method = "pairwise")

# veg height split by meadow type
m.time.veg.time <- glmmTMB(height.mean ~ treatment * meadow.type + 
                            (1|date) + (1|block/plot),
                  data = cattle.control)
summary(m.time.veg.time) # interaction not significant

# veg cover
m.time.veg <- glmmTMB(veg.2 ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.veg)
emm.time.veg <- emmeans(m.time.veg, specs = ~ treatment,
                            type = "response")
emm.time.veg
contrast(emm.time.veg, method = "pairwise")

# litter cover
m.time.litter <- glmmTMB(litter.2 ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.litter)
emm.time.litter <- emmeans(m.time.litter, specs = ~ treatment,
                            type = "response")
emm.time.litter
contrast(emm.time.litter, method = "pairwise")


# darlingtonia cover
m.time.dc <- glmmTMB(darlingtonia.2 ~ treatment + watershed + 
                       (1|date) + (1|block/plot),
                  data = filter(cattle.control, meadow.type == "wet"),
                  family = beta_family(link = "logit"))
emm.time.dc <- emmeans(m.time.dc, specs = ~ treatment)
emm.time.dc
contrast(emm.time.dc, method = "pairwise")

# water cover
m.time.water <- glmmTMB(water.2 ~ treatment + meadow.type + watershed + 
                            (1|date) + (1|block/plot),
                  data = cattle.control,
                  family = beta_family(link = "logit"))
summary(m.time.water )
emm.time.water  <- emmeans(m.time.water , specs = ~ treatment,
                            type = "response")
emm.time.water 
contrast(emm.time.water , method = "pairwise")

# manure count
m.time.manure <- glmmTMB(manure.count ~ treatment + meadow.type + watershed +
                            (1|date) + (1|block/plot),
                  data = filter(cattle.control, timing == "end of season"),
                  family = poisson(link = "log"))
summary(m.time.manure)
emm.time.manure <- emmeans(m.time.manure, specs = ~ treatment,
                            type = "response")
emm.time.manure
contrast(emm.time.manure, method = "pairwise")

# summary averages
cattle.sum <- cattle.control %>%
  group_by(treatment) %>%
  summarise(
    grazed.mean = mean(grazed.perc, na.rm = TRUE),
    trample.mean = mean(trample.both, na.rm = TRUE),
    height.mean = mean(height.mean, na.rm = TRUE),
    veg.mean = mean(veg.cover, na.rm = TRUE),
    litter.mean = mean(litter, na.rm = TRUE),
    dc.mean = mean(darlingtonia, na.rm = TRUE),
    manure.mean = mean(manure.count, na.rm = TRUE)
  )
cattle.sum

cattle.sum.2 <- cattle.control %>%
  filter(timing == "end of season") %>%
  group_by(treatment, year, watershed) %>%
  summarise(
    grazed.mean = mean(grazed.perc, na.rm = TRUE),
    trample.mean = mean(trample.both, na.rm = TRUE),
    height.mean = mean(height.mean, na.rm = TRUE),
    veg.mean = mean(veg.cover, na.rm = TRUE),
    litter.mean = mean(litter, na.rm = TRUE),
    dc.mean = mean(darlingtonia, na.rm = TRUE),
    manure.mean = mean(manure.count, na.rm = TRUE)
  )
View(cattle.sum.2)

```
In final year collected inflorescence abundance data
```{r flowering}
cattle.final <- cattle.ex %>% 
  filter(year == 2026) %>%
  select(block, treatment, timing, x1m_forb_infl, x1m_gram_infl,
         meadow.type, watershed) %>%
  mutate(timing = str_replace_all(timing," ", "_")) %>%
  pivot_wider(names_from = timing, values_from = c(x1m_forb_infl, x1m_gram_infl)) %>%
  mutate(forb_max = pmax(x1m_forb_infl_mid_season, x1m_forb_infl_end_of_season, 
                         na.rm = TRUE),
         gram_max = pmax(x1m_gram_infl_mid_season, x1m_gram_infl_end_of_season, 
                         na.rm = TRUE),
         infl_tot = forb_max + gram_max)


m.forb <- glmmTMB(forb_max ~ treatment + meadow.type + (1|block),
                  data = cattle.final,
                  family = poisson(link = "log"))
summary(m.forb)
emm.forb  <- emmeans(m.forb , specs = ~ treatment,
                            type = "response")
emm.forb
contrast(emm.forb , method = "pairwise")

m.gram <- glmmTMB(gram_max ~ treatment + meadow.type + (1|block),
                  data = cattle.final,
                  family = poisson(link = "log"))
summary(m.gram)
emm.gram  <- emmeans(m.gram , specs = ~ treatment,
                            type = "response")
emm.gram
contrast(emm.gram, method = "pairwise")

m.infl <- glmmTMB(infl_tot ~ treatment + meadow.type + 
                    (1|block),
                  data = cattle.final,
                  family = poisson(link = "log"))
summary(m.infl)
emm.infl  <- emmeans(m.infl , specs = ~ treatment,
                            type = "response")
emm.infl
contrast(emm.infl, method = "pairwise")

m.infl.2 <- glmmTMB(infl_tot ~ treatment + meadow.type + 
                    (1|block),
                  data = cattle.final)
summary(m.infl.2)
emm.infl.2  <- emmeans(m.infl.2 , specs = ~ treatment,
                            type = "response")
emm.infl.2
contrast(emm.infl.2, method = "pairwise")


cattle.final %>%
  ggplot(aes(x = treatment, y = infl_tot, 
             fill = treatment)) +
 geom_violin(alpha = 0.7, linewidth = 0, scale = "count", adjust = 2) +
  geom_point(position = position_jitter(width = 0.07, height = 1),
             alpha = 0.7)  +
  theme_bw() +
  scale_fill_manual(values = treat.cols.4) +
  ylab("Inflorescence count") +
  xlab("Treatment") +
  theme(legend.position = "none")

ggsave("Figures/cattle-exclosure_inf.pdf", width = 5, height = 4)

```


## Unused
Raw data visualizations
```{r visualizations-raw}

# trampling impact raw data
cattle.control %>% 
 # filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = animal.trample, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
#  geom_line(position = position_dodge(width = 50)) +
  theme_bw() 

# grazed percent raw data
cattle.control %>% 
  ggplot(aes(x = date, y = grazed.perc, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

# veg height raw data
cattle.control %>% 
  ggplot(aes(x = date, y = height.mean, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

# veg cover raw data
cattle.control %>% 
  ggplot(aes(x = date, y = veg.cover, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw()

# Dc cover raw data
cattle.control %>% 
  filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = darlingtonia, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  geom_line(position = position_dodge(width = 30)) +
  theme_bw() 

# water cover raw data
cattle.control %>% 
  ggplot(aes(x = date, y = water, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

```


Old analyses: starting with just looking at the final time point rather than a repeated measures structure
```{r old}

# filter dataset to just final time point
cattle.end <- cattle.control %>% filter(date == "2026-08-31")

# trampling (treatment impact within meadow type)
m.trample <- lmer(trample ~ treatment * meadow.type + (1|block), cattle.end)
emm.trample <- emmeans(m.trample, specs = ~ treatment | meadow.type)
emm.trample
contrast(emm.trample, method = "pairwise")

# trampling (averaged across meadow type)
m.trample <- lmer(trample ~ treatment + meadow.type + (1|block), cattle.end)
emm.trample <- emmeans(m.trample, specs = ~ treatment)
emm.trample
contrast(emm.trample, method = "pairwise")


# vegetation height 
m.height <- lmer(height.mean ~ treatment + meadow.type + (1|block), cattle.end)
emm.height <- emmeans(m.height, specs = ~ treatment )
emm.height
contrast(emm.height, method = "pairwise")


# vegetation cover 
m.veg <- lmer(veg.cover ~ treatment + meadow.type + (1|block), cattle.end)
emm.veg <- emmeans(m.veg, specs = ~ treatment)
emm.veg
contrast(emm.veg, method = "pairwise")

# Darlingtonia (wet meadows only)
cattle.end.wet <- cattle.end %>% filter(meadow.type == "wet")
m.dc <- lmer(darlingtonia ~ treatment + (1|block), cattle.end.wet)
emm.dc <- emmeans(m.dc, specs = ~ treatment)
emm.dc
contrast(emm.dc, method = "pairwise")


# Darlingtonia change
cattle.change.end <- cattle.change %>%
  filter(meadow.type == "wet", date > "2025-09-01")

m.dc.change <- lmer(dc.change ~ treatment + (1|block), cattle.change.end)
emm.dc.change <- emmeans(m.dc.change, specs = ~ treatment)
emm.dc.change
contrast(emm.dc.change, method = "pairwise")

```
