---
title: "SRWC-cattle-exclosures_analysis"
output: html_document
date: "2026-05-22"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(emmeans)

# list of wet meadow vs dry meadow blocks
wet.blocks <- c(2, 3, 5, 6, 8)

cattle.ex <- read.csv("data/SRWC-cattle-exclosure_post-treatment_2025-09-14.csv") %>%
  rename_with(tolower) %>%
  rename(manure.count = x..cowpies, 
         grazed.perc = x..grazed, 
         height.1 = height.1..cm.) %>%
  separate_wider_delim(plot, " ", names = c("position", "treatment")) %>%
  mutate(date = as.Date(date, "%m/%d/%y"),
         plot = paste0(block, position),
         height.mean = (height.1 + height.2 + height.3 + height.4)/4,
         meadow.type = case_when(block %in% wet.blocks ~ "wet", 
                                 TRUE ~ "dry"))



# compare versus control in each plot at each time step for height, veg.cover, 
# animal.trample, water, darlingtonia, grazed.perc
cattle.control <- cattle.ex %>%
  group_by(block, date) %>%
  mutate(
    height.control = height.mean[treatment == "control"][1],
    height.diff = height.mean - height.control,
    
    veg.control = veg.cover[treatment == "control"][1],
    veg.diff = veg.cover - veg.control,
    
    trample.control = animal.trample[treatment == "control"][1],
    trample.diff = animal.trample - trample.control,
    
    water.control = water[treatment == "control"][1],
    water.diff = water - water.control,
    
    dc.control = darlingtonia[treatment == "control"][1],
    dc.diff = darlingtonia - dc.control,
    
    grazed.control = grazed.perc[treatment == "control"][1],
    grazed.diff = grazed.perc - grazed.control
  ) %>%
  ungroup()

# compare to pre-treatment data
# cattle_change

  
```

## Scattershot visualizations
Currently for each block, comparing to the control using just a difference -- could also think about a proportion response or a log response ratio

Or we can try to compare first year's data to third year's data and look at what changed, but it's also interesting to see the difference in ex trampling effects. Also need to think about how we're managing the two time points of data each year


```{r visualizations-control}


# trampling impact compared to control
cattle.control %>% 
  ggplot(aes(x = date, y = trample.diff, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
#  geom_line(position = position_dodge(width = 50)) +
  theme_bw() 


# grazed percent compared to control
cattle.control %>%
  ggplot(aes(x = date, y = grazed.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 


# veg height compared to control
cattle.control %>%
  ggplot(aes(x = date, y = height.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 


# veg cover compared to control
cattle.control %>% 
  ggplot(aes(x = date, y = veg.diff, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw()


# Dc cover compared to control
cattle.control %>%
  ggplot(aes(x = date, y = dc.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

# Dc cover raw data change through time
cattle.control %>% 
  filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = darlingtonia, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  geom_line(position = position_dodge(width = 30)) +
  theme_bw() 


# water cover compared to control
cattle.control %>%
  ggplot(aes(x = date, y = water.diff, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 


```

# Preliminary analyses
Starting with just looking at the final time point rather than a repeated measures structure
```{r}

# filter dataset to just final time point
cattle.end <- cattle.control %>% filter(date > "2025-09-01")

# trampling (treamtment impact within meadow type)
m.trample <- lmer(animal.trample ~ treatment * meadow.type + (1|block), cattle.end)
emm.trample <- emmeans(m.trample, specs = ~ treatment | meadow.type)
emm.trample
contrast(emm.trample, method = "pairwise")

# trampling (averaged across meadow type)
m.trample <- lmer(animal.trample ~ treatment + meadow.type + (1|block), cattle.end)
emm.trample <- emmeans(m.trample, specs = ~ treatment)
emm.trample
contrast(emm.trample, method = "pairwise")

```



## Unused
Raw data visualizations
```{r visualizations-raw}

# trampling impact raw data
cattle.control %>% 
 # filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = animal.trample, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
#  geom_line(position = position_dodge(width = 50)) +
  theme_bw() 

# grazed percent raw data
cattle.control %>% 
  ggplot(aes(x = date, y = grazed.perc, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

# veg height raw data
cattle.control %>% 
  ggplot(aes(x = date, y = height.mean, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

# veg cover raw data
cattle.control %>% 
  ggplot(aes(x = date, y = veg.cover, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw()

# Dc cover raw data
cattle.control %>% 
  filter(month(date) == 9) %>%
  ggplot(aes(x = date, y = darlingtonia, 
             color = treatment, shape = meadow.type, group = plot)) +
  geom_point(position = position_dodge(width = 30)) +
  geom_line(position = position_dodge(width = 30)) +
  theme_bw() 

# water cover raw data
cattle.control %>% 
  ggplot(aes(x = date, y = water, 
             color = treatment, shape = meadow.type)) +
  geom_point(position = position_dodge(width = 30)) +
  theme_bw() 

```

