---
title: "SRWC-seedbank_greenhouse_prelim-analysis"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)

# Color Options:

color.palatte <- c("mediumblue", "darkred") 
fun.palatte <- c("darkolivegreen","darkgoldenrod")
```

## Read in Data

Two csv files for greenhouse study data, one from 2023 start sampling and one from 2024 start sampling. Each of these files has a corresponding plant ID file and they do still need to be cleaned up and converted to species codes

```{r read-files}

# year 1 greenhouse data
greenhouse.n <- read.csv(here("data/SRWC-seedbank_greenhouse_2024-05-22_n.csv"), 
                         header = TRUE)

greenhouse.n$X <- NULL
greenhouse.n$habitat.type <- NULL

greenhouse.n$species <- greenhouse.n$species %>% 
  trimws() %>% tolower() %>%
  str_replace_all(" ", ".")

# switch to long format (year 1 data)
greenhouse.long <- greenhouse.n %>% 
  pivot_longer(cols = rf.1.a:control, names_to = "plot", values_to = "count",
               values_drop_na = TRUE) %>%
  mutate(sample.year = 2023)



# year 2 greenhouse data
greenhouse.2 <- read.csv(here("data/SRWC-seedbank_greenhouse_year-2_2025-05-27.csv"), 
                         header = TRUE) %>%
  rename_with(tolower)

greenhouse.2.long <- greenhouse.2 %>%
  mutate(across(rf.1.a:control, as.integer),
         species = case_when(species.code > 0 ~ species.code, 
                             .default = species.morpho)) %>%
  pivot_longer(cols = rf.1.a:control, names_to = "plot", values_to = "count",
               values_drop_na = TRUE) %>%
  mutate(sample.year = 2024)

greenhouse.2.long$species <- greenhouse.2.long$species %>% 
  trimws() %>% tolower() %>%
  str_replace_all(" ", ".") 

# join year 1 and year 2 greenhouse data (long format)
# some data cleaning as needed for names with errors
greenhouse <- full_join(greenhouse.2.long, greenhouse.long, 
                        by = c("sample.year", "plot", "species", "count",
                               "date", "observer")) %>%
  mutate(  
    species = case_when(
      species == "branching.long.leaf" ~ "sticky.grass",
      species == "probably.viola" ~ "vioadu",
      species == "unidentified.sedge?" ~ "junsp?",
      .default = species
    )
  )

```



## Summarize for max species abundance

summarize across dates for max abundance of each species in each plot
```{r summarize-date}

# summary by species max count
greenhouse.sum <- greenhouse %>% 
  group_by(species, plot, sample.year) %>% 
  dplyr::summarise(
    count.max = max(count, na.rm = TRUE)) %>%
  filter(count.max > 0)
  

# View(greenhouse.sum)

# remove species present in control trays
control.species <- greenhouse.sum %>% 
  ungroup() %>%
  filter(plot == "control") %>% 
  add_row(species = "jagged.fred") %>%
  pull(species) 

greenhouse.sum.clean <- greenhouse.sum %>%
  filter(!(species %in% control.species), !(species == "butterfly.guy")) %>%
  rename(greenhouse.count.max = "count.max") %>%
  separate_wider_delim(plot, delim = ".", names = c("meadow", "plot.id", "habitat"),
                       cols_remove = FALSE) %>%
  arrange(meadow, habitat, plot.id, species)

write.csv(greenhouse.sum.clean, "SWRC-seedbank_greenhouse_summary.csv")
```

# Data Cleaning
TO DO: Updating morpho names for species names where they were recently keyed, adding in observations of ones that were put into bigger pots but not recorded on the weekly sheets, and sorting out an error where three different species were initially conflated when young

```{r greenhouse-data-cleaning}
greenhouse.edit <- greenhouse.sum.clean %>%
  mutate(
    species = case_when(
      
      
      .default = species
    )
  )



```



# Field data
Read it in, organize and summarize it in the same way to match the greenhouse.max.summary dataset. Combine field and greenhouse data
```{r read-data-2}

field.data.raw <- read.csv(here("data/SRWC-seedbank_plant-survey_2023-07-21b.csv"), header = TRUE)

# removing duplicate rows
dup.1 <- which(field.data.raw$plot == "rf.6.a" & field.data.raw$species.code == "unkfb")[2]
dup.2 <- which(field.data.raw$plot == "rf.4.p" & field.data.raw$species.code == "sidore")[2]
drop <- c(dup.1, dup.2)

field.data <- field.data.raw %>% 
  filter(!row_number() %in% drop) %>%
  rename(field.present.2023 = "species.count")


```

# Combine field and greenhouse data
Right now just using 2023 field data, also have 2024 data with additional species and frequency estimates
```{r combine-field-greenhouse}
meadows.data <- field.data %>% 
  full_join(greenhouse.sum.clean, 
            by = c("plot","meadow", "habitat", "species.code"="species")) 

meadows.species.sum <- meadows.data %>%
  group_by(meadow, habitat, species.code) %>%
  summarise(
    plots.present.field = sum(field.present.2023 > 0, na.rm = TRUE),
    plots.present.greenhouse = sum(greenhouse.count.max > 0, na.rm = TRUE)
  ) 

meadows.species.wide <- meadows.species.sum %>%
  pivot_wider(
    names_from = c(meadow, habitat),
    values_from = c(plots.present.field, plots.present.greenhouse),
    names_vary = "slowest"
  ) %>%
  mutate(
    across(everything(), ~replace_na(.x,0))
  ) %>%
  sort(species.code)

write.csv(meadows.species.wide, "SRWC-seedbank_species-frequency.csv")


```





# Old analyses from SOU 2023 Capstone -----------------------

# Summarize greenhouse and field data for richness

```{r summarize-data-2}

greenhouse.max.summary <- greenhouse.sum.clean %>%
  group_by(plot) %>%
  dplyr::summarise(
    greenhouse.richness = sum(count.max > 0, na.rm = TRUE)
  )

greenhouse.max.summary


field.summary <- field.data %>%
  group_by(plot, habitat) %>%
  dplyr::summarise(
    field.richness = sum(species.count > 0, na.rm = TRUE))

field.summary

```

# Combine field and greenhouse data
Use the function left_join(dataframe1, dataframe2, by = "plot")
```{r join-data-1}

combined.sum <- left_join(greenhouse.max.summary, field.summary, by = "plot")

```

# Omit missing data (control)
```{r omit-data-1}

combined.summary <- na.omit(combined.sum)

```


# Begin linear regression:
```{r regress-data-1}

combined.lm = lm(greenhouse.richness ~ field.richness, data = combined.summary)

combined.lm

summary(combined.lm)
```
# Begin linear regression:
```{r graph-regression-1}

GH.vs.field.regression <- ggplot(combined.summary, aes(x=field.richness, y=greenhouse.richness)) +
    geom_point()+
    theme_classic()+
    xlab("Field Richness")+
    ylab("Greehouse Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

GH.vs.field.regression

ggsave(plot = GH.vs.field.regression, filename = "figures/GH-vs-field-regression.png", width = 6, height = 4, units = "in")

```
# Do an anova to compare habitat type:

# Field - Looking at distribution:
```{r anova-1}

boxplot(field.richness~habitat, data=combined.summary, main="Field Richness by Habitat Type", xlab = "Habitat", ylab = "Species Richness")
```
```{r ggboxplot-1}
  
field.richness.by.habitat <- combined.summary %>%
      ggplot(aes(x=habitat, y=field.richness))+
      geom_boxplot(aes(fill=habitat, col = fun.palatte), colour="black") +
      labs(x= "Habitat Type", y="Field Species Richness")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow", "N/A")) +
        scale_y_continuous(limits=c(5,25), breaks=c(5, 10, 15, 20, 25))+
        scale_fill_manual(values= fun.palatte) +
  theme(legend.position = "none")

field.richness.by.habitat

ggsave(plot = field.richness.by.habitat, filename = "figures/field_richness_by_habitat.png", width = 6, height = 4, units = "in")
```

#Field ANOVA time:
```{r anova-4}

fieldrichness.aov = aov(field.richness~habitat, data=combined.summary)

  summary(fieldrichness.aov)
  
```
# Greenhouse - Looking at distribution:
```{r anova-5}

boxplot(greenhouse.richness~habitat, data=combined.summary, main="Greenhouse Richness by Habitat Type", xlab = "Habitat", ylab = "Species Richness") 
 
```
#Greehouse ANOVA time:
```{r ggboxplot-2}
  
GH.richness.by.habitat <- combined.summary %>%
      ggplot(aes(x=habitat, y=greenhouse.richness))+
      geom_boxplot(aes(fill=habitat, col = color.palatte), colour="black") +
      labs(x= "Habitat Type", y="Greenhouse Species Richness")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow", "N/A")) +
        scale_y_continuous(limits=c(0,15), breaks=c(0, 5, 10, 15))+
        scale_fill_manual(values= fun.palatte)+
        theme(legend.position = "none")

GH.richness.by.habitat

ggsave(plot = GH.richness.by.habitat, filename = "figures/GH_richness_by_habitat.png", width = 6, height = 4, units = "in")

```

#Greehouse ANOVA time:

```{r anova-6}

greenhouserichness.aov = aov(greenhouse.richness~habitat, data=combined.summary)

  summary(greenhouserichness.aov)
  
```
#Two way ANOVA?:
```{r anova-7}

habitat.richness = aov(field.richness ~ habitat*greenhouse.richness, data=combined.summary)

summary(habitat.richness)
  
```
#Soil moisture 
```{r sm-read data}

##soil.moisture <- SRWC_seedbank_soil_moisture_2023_09_20

soil.moisture <- read.csv(here("data/SRWC-seedbank_soil-moisture_2023-09-20.csv"), header = TRUE)

combined.summary <- left_join(combined.summary, soil.moisture, by = c("plot","habitat"))
  
```

#soil moisture summary 

```{r SM-summarize-data}

moisture.lm = lm(moisture.content ~ habitat, data = soil.moisture)

moisture.lm

summary(moisture.lm)

```
#soil moisture ANOVA boxplot 
```{r SM-ggboxplot-1}
  
soil.moisture.plot <- soil.moisture %>%
      ggplot(aes(x=habitat, y=moisture.content))+
      geom_boxplot(aes(fill=habitat, col = color.palatte), colour="black") +
      labs(x= "Habitat Type", y="Soil Moisture Content (%)")+
      theme_classic() +
        scale_x_discrete(labels=c("Actual Meadow", "Potential Meadow")) +
        scale_y_continuous(limits=c(0,100), breaks=c(0, 25, 50, 75, 100))+
        scale_fill_manual(values= fun.palatte)+
        theme(legend.position = "none")

soil.moisture.plot

ggsave(plot = soil.moisture.plot, filename = "figures/soil-moisture-plot.png", width = 6, height = 4, units = "in")
    
```

#soil moisture ANOVA time!! 

```{r anova-8}

soilmoisture.aov = aov(moisture.content~habitat, data=soil.moisture)

  summary(soilmoisture.aov)
  
```
```{r SM-regress-data}

SMfield.lm = lm(field.richness ~ moisture.content, data=combined.summary)

SMfield.lm

summary(SMfield.lm)

```
```{r SM-regress-data-2}

SMGH.lm = lm(greenhouse.richness ~ moisture.content, data=combined.summary)

SMGH.lm

summary(SMGH.lm)

```
```{r anova-9}

habitat.richness = aov(field.richness ~ habitat*moisture.content, data=combined.summary)

summary(habitat.richness)

habitat.richness.2 = aov(greenhouse.richness ~ habitat*moisture.content, data=combined.summary)

summary(habitat.richness.2)
  
```

```{r anova-10}

habitat.names <- c(a="Actual Meadow", p="Potential Meadow")

field.richness.vs.soil.moisture <- ggplot(combined.summary, aes(x=moisture.content, y=field.richness)) +
  facet_wrap(vars(habitat),labeller=labeller(habitat=habitat.names), scales="free_x")+
    geom_point()+
    theme_classic()+
    xlab("Soil Moisture Content (%)")+
    ylab("Field Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

field.richness.vs.soil.moisture

ggsave(plot = field.richness.vs.soil.moisture, filename = "figures/field_richness_vs_soil_moisture.png", width = 6, height = 4, units = "in")

GH.richness.vs.soil.moisture <- ggplot(combined.summary, aes(x=moisture.content, y=greenhouse.richness)) +
  facet_wrap(vars(habitat),labeller=labeller(habitat=habitat.names), scales="free_x")+
    geom_point()+
    theme_classic()+
    xlab("Soil Moisture Content (%)")+
    ylab("Greenhouse Richness")+
         geom_smooth(method=lm, color="darkgreen", fill="olivedrab")

GH.richness.vs.soil.moisture

ggsave(plot = GH.richness.vs.soil.moisture, filename = "figures/GH_richness_vs_soil_moisture.png", width = 6, height = 4, units = "in")


```